---
title: 'Bios 6301: Assignment 3'
author: "Josh DeClercq"
output: pdf_document

---

*Due Tuesday, 11 October, 1:00 PM*

50 points total.

$5^{n=day}$ points taken off for each day late.

This assigment includes turning in the first two assignments. All three should
include knitr files (named `homework1.rmd`, `homework2.rmd`, `homework3.rmd`)
along with valid PDF output files. Inside each file, clearly indicate which
parts of your responses go with which problems (you may use the original homework
document as a template). Add your name as `author` to the file's metadata section.
Raw R code/output or word processor files are not acceptable.

Failure to properly name files or include author name may result in 5 points taken off.

### Question 1 ###

**10 points**

1. Use GitHub to turn in the first three homework assignments. Make sure the teacher (couthcommander) and TA (chipmanj) are collaborators. (5 points)

1. Commit each assignment individually.  This means your repository should have at least three commits. (5 points)

### Question 2 ###

**15 points**

Write a simulation to calculate the power for the following study
design.  The study has two variables, treatment group and outcome.
There are two treatment groups (0, 1) and they should be assigned
randomly with equal probability.  The outcome should be a random normal
variable with a mean of 60 and standard deviation of 20.  If a patient
is in the treatment group, add 5 to the outcome.  5 is the true
treatment effect.  Create a linear model for the outcome by the
treatment group, and extract the p-value (hint: see assignment1).
Test if the p-value is less than or equal to the alpha level, which
should be set to 0.05.

Repeat this procedure 1000 times. The power is calculated by finding
the percentage of times the p-value is less than or equal to the alpha
level.  Use the `set.seed` command so that the professor can reproduce
your results.

```{r cache = TRUE, tidy = TRUE}
getpower = function(patients, nsim) {
  p.test <- numeric(nsim)
  set.seed(12356)
  for(i in seq_along(p.test)) {
    outcome = rnorm(patients, 60, 20)
    treatment <- sample(patients)%%2 
    p.test[i] = summary(lm(outcome ~ treatment))$coefficients[2,4]
  }
  mean(p.test < 0.05)
}
ans1 <- getpower( patients = 100, nsim = 1000)
ans2 <- getpower( patients = 1000, nsim = 1000)
true.power <- power.t.test(n=100, delta=0.5, sd=20, sig.level=0.05, type='one.sample')$power
```

1. Find the power when the sample size is 100 patients. (10 points)

*The power when the sample size is 100 patients is `r ans1`.*

2. Find the power when the sample size is 1000 patients. (5 points)

*The power when the sample size is 1000 patients is `r ans2`. The true power for this test is `r true.power`.*

### Question 3 ###

**15 points**

Obtain a copy of the [football-values lecture](https://github.com/couthcommander/football-values).
Save the `2016/proj_wr16.csv` file in your working directory.  Read
in the data set and remove the first two columns.

1. Show the correlation matrix of this data set. (3 points)
```{r}
my.df <- read.csv('/Users/Bowie/Dropbox/Stat_Com/proj_wr16.csv')
```

```{r}
library(abind)
library(corrplot)
library(MASS)
```

```{r}
head(my.df)
my.df <- my.df[ -c(1, 2) ]

head(my.df)
df.cor <- cor(my.df)
df.cor
corrplot(df.cor, method = "circle")
length(my.df$rush_att)
```
2. Generate a data set with 30 rows that has a similar correlation
structure.  Repeat the procedure 10,000 times and return the mean
correlation matrix. (10 points)

*I really struggled with this one, and I had to download a new package (abind) to get it to work.*
```{r cache = TRUE, warning = FALSE}

all.df <- list(replicate(10000, cor(my.df[sample(243,30,replace = FALSE),])))
all.matrix <- abind(all.df, along=3)
M <- apply(all.matrix, c(1,2),mean, na.rm = TRUE)
M
corrplot(M, method = "circle")
```

3. Generate a data set with 30 rows that has the exact correlation
structure as the original data set. (2 points)

*I was not sure what is meant by 'exact correlation structure.' Here, I attempted to find 30 entries of the dataframe that make a correlation structure exactly equal to the correlation structure of all 243 entries. I think that this may be impossible to do.*
```{r cache = TRUE, warning = FALSE, eval = FALSE}
x<- 0
repeat{
  x <- cor(my.df[sample(243,30,replace = FALSE),])
  if( all.equal(df.cor, x, tolerance = 0.01) == TRUE) break
}
x
```

*I attempted to use some code I had in the lecture notes to achieve the same result. It works, but have to set a tolerance threshold in the test for equality.*
```{r cache = TRUE, warning = FALSE}
vcov.df <- var(my.df)
means.df <- colMeans(my.df)

keep.1 <- 0
loops <- 1e4
for (i in seq(loops)){
df.sim <- mvrnorm(30, mu = means.df, Sigma = vcov.df, empirical = TRUE)
keep.1 <- keep.1 + cor(df.sim)/loops
}
keep.1
 all.equal(df.cor, keep.1, tolerance = 0.0000000001)
```

*Below are two attmpts I made to extract the 30 rows of data that would generate the correlation matrix, but I definitely did not figure this one out. When we went over simulation in lecture, I had a really hard time keeping up with copying the code and I hardly had a chance to understand what was happening nor did I have a chance to annotate my code so I could make sense of it later. I could not find anything like this in the class notes or github repository to help me along.* 
```{r cache = TRUE, warning = FALSE, eval = FALSE}
vcov.df <- var(my.df)
means.df <- colMeans(my.df)

df.sim <- list()
keep.1 <- 0
loops <- 1e4
for (i in seq(loops)){
df.sim[i] <- mvrnorm(30, mu = means.df, Sigma = vcov.df, empirical = TRUE)/loops
}
X.matrix <- abind(df.sim, along=3)
X <- apply(X.matrix, c(1,2),mean, na.rm = TRUE)
X
cor(X)
```

```{r cache = TRUE, warning = FALSE, eval = FALSE, tidy = TRUE}
X.df <- list(replicate(10000, mvrnorm(30, mu = means.df, Sigma = vcov.df, 
                                      empirical = TRUE)))
X.matrix <- abind(X.df, along=3)
X <- apply(X.matrix, c(1,2),mean, na.rm = TRUE)
X
cor(X)
```
### Question 4 ###

**10 points**

Use LaTeX to create the following expressions.  
*Note: These look better in html than pdf. I couldn't figure out how to make the \large functions work with the pdf output. Also, is it possible to load Latex packages into Rmarkdown?*

1. Hint: `\Rightarrow` (4 points)

$$\LARGE{P(B) = \sum_{j} P(B | A_j) P(A_j),\\
\Rightarrow P(A_i|B) = } \frac{P(B | A_i) P(A_i)} {\sum_j P(B|A_j) P(A_j)}$$

2. Hint: `\zeta` (3 points)

 $$\LARGE{\hat{f}(\zeta) = \displaystyle\int_{-\infty}^{\infty} f(x)e^{-2\pi ix \zeta} dx}$$

3. Hint: `\partial` (3 points)

$$\LARGE{\textbf{J} = \frac{d\textbf{f}}{d\textbf{x}} = 
\begin{bmatrix} 
\frac{\partial \textbf{f}}{\partial x_1}  \dots  \frac{\partial \textbf{f}}{\partial x_n} \\
\end{bmatrix} = 
\begin{bmatrix} 
\frac{\partial f_1}{\partial x_1} & \dots & \frac{\partial f_1}{\partial x_n} \\
\vdots & \ddots & \vdots \\
\frac{\partial f_m}{\partial x_1} & \dots & \frac{\partial f_m}{\partial x_n} \\
\end{bmatrix}}
$$
